Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-21 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
First draft of data section

@@

\noindent and therefore:
$$
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
$$

@@

>>>>

# CRediT author statement

**Author 1:** Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Visualization, Supervision; **Author 2:** Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Visualization; **Author 3:** Conceptualization, Investigation, Data Curation, Writing - Original Draft; **Author 4:** Resources - Writing: Review & Editing - Supervision

@@

\noindent and also:
$$
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
$$

@@

Data have been prepared for sharing as a data package^[add repository].

### Statistics Canada

Population and income statistics at the level of Dissemination Areas (DAs) were retrieved using the package `cancensus` [@vonBergmann2021cancensus]. Dissemination Areas are the smallest publicly available census geography in Canada. We use data from the 2016 Population Census.

### Origins: Residential parcels

We converted all recorded residential parcels in the City of Hamilton to points on the road network. Each point includes information about the number of residential units in the parcel. Each residential unit is "populated" with the probability of being a "low income household", which we define as having a total household income of less than CAD 40,000. This is approximately the mid-point of the low income cut-off (LICOs) for families in Canadian cities with populations greater than 500,000 in 2016, to match other Census data [@statisticscanada2020licos].

### Destinations: Foodbanks

**The locations of foodbanks were obtained from public records and geocoded. Three urban sites not in the program were also identified and geocoded for comparison purposes.**

### Routing and travel time tables

Travel time tables for three modes (car, transit, walking) were computed using the parcels as the origins and the locations of the foodbanks as the destinations. For routing, the package `r5r` [@pereira2021r5r] was used, with a network extract for the City of Hamilton from OpenStreetMaps and the General Transit Feed Specification from Hamilton Street Railway, the local transit operator. For routing purposes we used maximum values of 180 min and 10,000 m walking distance: any destination that exceeded these thresholds was ignored. The departure time used for routing was **TIME**.

### Open Hamilton **???**

We used the open data portal of the City of Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained boundaries for the city's various regions (the definition of urban, suburban, and rural regions follows the classification of development applications).

### Transportation Tomorrow Survey

We used the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}} to download cross-tabulations of: 1) primary mode of travel per trip by income by place of residence; and 2) age by income by place of residence. These data are from the 2016 Survey (the most recent available), and data are geocoded at the level of Traffic Analysis Zones (TAZ) using the most recent zoning system from 2006. Each parcel point is populated with the proportion of trips by three modes of travel: car (as driver or passenger), transit, and walk.

### Expected Travel Times

Once we obtained travel time tables with population (number of households) and proportion of trips by mode, we calculated the expected travel time $ett$ from each parcel $i$ to a foodbank $j$ as follows:
$$
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the foodbank. In other words, the expected travel time is the weighted sum of travel times to the foodbank, with the weights given by the expected modal split in the TAZ.

### 2021-04-21 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
First draft of data section

@@

But the emergence of COVID-19, the worst public health crisis since the
1918 flu pandemic, has revealed important social and economic fault
lines, and pre-existing patterns of inequality appear to have been
exacerbated. Along several other dimensions of stress {[}e.g.,
accessibility to health care facilities; Pereira et al. (2021){]}, this
seems to be the case for food insecurity as well (Laborde et al., 2020).
In the US, for example, it has been estimated that there was an increase
of more than 30\% in household food insecurity, and more than one third
of households were discovered to be newly food insecure - meaning they
did not experience food insecurity before the pandemic (Niles et al.,
2020). In Canada, Men and Tarasuk (2021) report that about 25\% of
individuals who experienced job insecurity (a relatively common
occurrence during the pandemic), also experienced food insecurity.
Similarly, according to Statistics Canada (2020), in the early stages of
the pandemic almost 15\% of individuals reported living in a household
that faced food insecurity; the risk of food insecurity was
substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children); likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

@@

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility.

@@

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). In 2012, 12.4\% of Canadians households and 11.8\% Ontarian
households experienced some degree of food insecurity (Gundersen et al.,
2018; Tarasuk et al., 2020). Pandemic-related food security studies in
the US have found a substantial increase in households experiencing food
insecurity for the first time, and also in households experiencing more
severe food insecurity than before (Niles et al., 2020; Wolfson and
Leung, 2020). Most recently in May 2020, Canada recorded 14.7\% of its
population living in food insecurity in the past 30 days (Statistics
Canada, 2020). Food insecurity is a highly concerning public health
issue due to its health impacts. Recent research suggests that food
insecurity in adults can lead to experiencing more stressful events
(El-hajj and Benhin 2021). Increases in food insecurity rates due to the
pandemic, signal a change in the food environment with potential damages
to population health during the course of the pandemic and beyond (Niles
et al., 2020).

@@

\noindent and therefore: \[
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
\]

@@

@@

\leavevmode\hypertarget{ref-pereira2021geographic}{}%
Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
Gouveia, N., Paez, A., 2021. Geographic access to COVID-19 healthcare in
brazil using a balanced float catchment area approach. Social Science \&
Medicine 273, 113773.
doi:\url{https://doi.org/10.1016/j.socscimed.2021.113773}

@@

\leavevmode\hypertarget{ref-statisticscanada2020food}{}%
Statistics Canada, 2020. Food insecurity during the COVID-19 pandemic
(No. Catalogue no. 45280001).

@@

@@

>>>>

\hypertarget{credit-author-statement}{%
\section{CRediT author statement}\label{credit-author-statement}}

\textbf{Author 1:} Conceptualization, Methodology, Software, Validation,
Formal analysis, Investigation, Data Curation, Writing - Original Draft,
Visualization, Supervision; \textbf{Author 2:} Conceptualization,
Methodology, Software, Validation, Formal analysis, Investigation, Data
Curation, Writing - Original Draft, Visualization; \textbf{Author 3:}
Conceptualization, Investigation, Data Curation, Writing - Original
Draft; \textbf{Author 4:} Resources - Writing: Review \& Editing -
Supervision

@@

But the emergence of COVID-19, the worst public health crisis since the
1918 flu pandemic, has revealed important social and economic fault
lines, and pre-existing patterns of inequality appear to have been
exacerbated. Along several other dimensions of stress {[}e.g.,
accessibility to health care facilities; Pereira et al. (2021a){]}, this
seems to be the case for food insecurity as well (Laborde et al., 2020).
In the US, for example, it has been estimated that there was an increase
of more than 30\% in household food insecurity, and more than one third
of households were discovered to be newly food insecure - meaning they
did not experience food insecurity before the pandemic (Niles et al.,
2020). In Canada, Men and Tarasuk (2021) report that about 25\% of
individuals who experienced job insecurity (a relatively common
occurrence during the pandemic), also experienced food insecurity.
Similarly, according to Statistics Canada (2020a), in the early stages
of the pandemic almost 15\% of individuals reported living in a
household that faced food insecurity; the risk of food insecurity was
substantially higher for households with children. The difference
between households with and without children was significant, and 11.7\%
of households with children indicated that ``food didn't last and
{[}there was{]} no money to get more'' sometimes or often, compared to
7.3\% of households without children); likewise, 13\% of households with
children indicated that they ``{[}c{]}ouldn't afford balanced meals''
sometimes or often, compared to 8.8\% of households without children.

@@

It is known that the urban food environment, within which people make
their daily food choices, is essential in influencing eating behaviours
and health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility. \textbf{Check
Tarasuk et al. (2019)}

@@

Currently with the COVID-19 pandemic, disrupted economies, rising
unemployment rates, and changes in poverty levels have disrupted the
food environment and led to higher rates of food insecurity (Niles et
al., 2020). In 2012, 12.4\% of Canadians households and 11.8\% Ontarian
households experienced some degree of food insecurity (Gundersen et al.,
2018; Tarasuk et al., 2020). Pandemic-related food security studies in
the US have found a substantial increase in households experiencing food
insecurity for the first time, and also in households experiencing more
severe food insecurity than before (Niles et al., 2020; Wolfson and
Leung, 2020). Most recently in May 2020, Canada recorded 14.7\% of its
population living in food insecurity in the past 30 days (Statistics
Canada, 2020a). Food insecurity is a highly concerning public health
issue due to its health impacts. Recent research suggests that food
insecurity in adults can lead to experiencing more stressful events
(El-hajj and Benhin 2021). Increases in food insecurity rates due to the
pandemic, signal a change in the food environment with potential damages
to population health during the course of the pandemic and beyond (Niles
et al., 2020).

@@

\noindent and also: \[
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
\]

@@

Data have been prepared for sharing as a data package\footnote{add
  repository}.

\hypertarget{statistics-canada}{%
\subsubsection{Statistics Canada}\label{statistics-canada}}

Population and income statistics at the level of Dissemination Areas
(DAs) were retrieved using the package \texttt{cancensus} (von Bergmann
et al., 2021). Dissemination Areas are the smallest publicly available
census geography in Canada. We use data from the 2016 Population Census.

\hypertarget{origins-residential-parcels}{%
\subsubsection{Origins: Residential
parcels}\label{origins-residential-parcels}}

We converted all recorded residential parcels in the City of Hamilton to
points on the road network. Each point includes information about the
number of residential units in the parcel. Each residential unit is
``populated'' with the probability of being a ``low income household,''
which we define as having a total household income of less than CAD
40,000. This is approximately the mid-point of the low income cut-off
(LICOs) for families in Canadian cities with populations greater than
500,000 in 2016, to match other Census data (Statistics Canada, 2020b).

\hypertarget{destinations-foodbanks}{%
\subsubsection{Destinations: Foodbanks}\label{destinations-foodbanks}}

\textbf{The locations of foodbanks were obtained from public records and
geocoded. Three urban sites not in the program were also identified and
geocoded for comparison purposes.}

\hypertarget{routing-and-travel-time-tables}{%
\subsubsection{Routing and travel time
tables}\label{routing-and-travel-time-tables}}

Travel time tables for three modes (car, transit, walking) were computed
using the parcels as the origins and the locations of the foodbanks as
the destinations. For routing, the package \texttt{r5r} (Pereira et al.,
2021b) was used, with a network extract for the City of Hamilton from
OpenStreetMaps and the General Transit Feed Specification from Hamilton
Street Railway, the local transit operator. For routing purposes we used
maximum values of 180 min and 10,000 m walking distance: any destination
that exceeded these thresholds was ignored. The departure time used for
routing was \textbf{TIME}.

\hypertarget{open-hamilton}{%
\subsubsection{\texorpdfstring{Open Hamilton
\textbf{???}}{Open Hamilton ???}}\label{open-hamilton}}

We used the open data portal of the City of
Hamilton\footnote{\url{https://open.hamilton.ca/}} we obtained
boundaries for the city's various regions (the definition of urban,
suburban, and rural regions follows the classification of development
applications).

\hypertarget{transportation-tomorrow-survey}{%
\subsubsection{Transportation Tomorrow
Survey}\label{transportation-tomorrow-survey}}

We used the Data Retrieval System of the Transportation Tomorrow Survey
(TTS)\footnote{\url{http://dmg.utoronto.ca/}} to download
cross-tabulations of: 1) primary mode of travel per trip by income by
place of residence; and 2) age by income by place of residence. These
data are from the 2016 Survey (the most recent available), and data are
geocoded at the level of Traffic Analysis Zones (TAZ) using the most
recent zoning system from 2006. Each parcel point is populated with the
proportion of trips by three modes of travel: car (as driver or
passenger), transit, and walk.

\hypertarget{expected-travel-times}{%
\subsubsection{Expected Travel Times}\label{expected-travel-times}}

Once we obtained travel time tables with population (number of
households) and proportion of trips by mode, we calculated the expected
travel time \(ett\) from each parcel \(i\) to a foodbank \(j\) as
follows: \[
ett_i = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
\] \noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the foodbank. In other words, the expected travel time
is the weighted sum of travel times to the foodbank, with the weights
given by the expected modal split in the TAZ.

@@

\leavevmode\hypertarget{ref-pereira2021geographic}{}%
Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
Gouveia, N., Paez, A., 2021a. Geographic access to COVID-19 healthcare
in brazil using a balanced float catchment area approach. Social Science
\& Medicine 273, 113773.
doi:\url{https://doi.org/10.1016/j.socscimed.2021.113773}

\leavevmode\hypertarget{ref-pereira2021r5r}{}%
Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga, C.K.V., Conway,
M.W., 2021b. r5r: Rapid realistic routing on multimodal transport
networks with r\textsuperscript{5} in r. Findings.
doi:\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}

@@

\leavevmode\hypertarget{ref-statisticscanada2020food}{}%
Statistics Canada, 2020a. Food insecurity during the COVID-19 pandemic
(No. Catalogue no. 45280001).

\leavevmode\hypertarget{ref-statisticscanada2020licos}{}%
Statistics Canada, 2020b. Table 11-10-0241-01 low income cut-offs
(LICOs) before and after tax by community size and family size, in
current dollars. doi:\url{https://doi.org/10.25318/1110024101-eng}

@@

\leavevmode\hypertarget{ref-tarasuk2019geographic}{}%
Tarasuk, V., Fafard St-Germain, A.-A., Mitchell, A., 2019. Geographic
and socio-demographic predictors of household food insecurity in canada,
2011--12. BMC Public Health 19, 12.
doi:\href{https://doi.org/10.1186/s12889-018-6344-2}{10.1186/s12889-018-6344-2}

@@

\leavevmode\hypertarget{ref-vonBergmann2021cancensus}{}%
von Bergmann, J., Shkolnik, D., Jacobs, A., 2021. Cancensus: R package
to access, retrieve, and work with canadian census data and geography.
